Legacy Data Confounds Genomics Studies

### 2019-07-09 LatexManuscript/legacy.tex
update fig text
minor text, pre meeting push

\begin{figure}
\scalebox{0.85}{\includegraphics[width=\hsize,keepaspectratio]{../Figures/Figure1.jpg}}
\caption{
Suspicious mutations carried by individuals with low quality data have distinct mutational profiles, reproduce poorly across studies, and are distributed across the genome.
\textbf{A} 
Genome wide association of the average quality of mapped bases $Q$ for the 104 Japanese individuals included in the 1000 Genomes Project. 
This GWAS identified $1034$ SNPs associated to the average $Q$ of SNPs mapped for an individual, with  $p < 10^{-6}$
($587$ SNPs had  $p < 10^{-8}.$)
\textbf{B} 
Joint frequency spectrum plot of the Japanese from the 1000 Genomes Project and a more recent Japanese dataset from Nagahama.
The size of blue dots are proportional to the number of variants in a given frequency bin that associate with $Q$ in the JPT. 
\textbf{C} 
Mutation spectrum of the 1034 variants that associated with $Q$ in the JPT($p < 10^{-6}$), compared to the expectation from the distribution of all SNPs.
The majority of the variants with significant associations to $Q$ have the *AC${\rightarrow}$*CC mutational pattern. 
There is also an enrichment in GA*${\rightarrow}$GG* and GC*${\rightarrow}$GG* mutations. 
These three enrichments can be summarized as G**${\rightarrow}$GG*. 
Stars ( * ) indicate a significant deviation from the expected mutational spectrum from all SNPs.
}
 \label{SFS}
\end{figure}

@@

These associations are not necessarily incorrect -- a weak but significant bias in imputation may still result in a correct associations.
To distinguish between variants with weak but significant association with $Q$ from variants with strong biases, we distinguished between variants where the allele frequency difference between individuals with low- and high-$Q$ is larger than a factor of two (which naturally separates two clusters of variants on Figure \ref{90HanSFS_full}).
The majority (92.7\%) of the $Q$-associated variants are strongly biased in that they are more than twice as frequent in individuals with low-$Q$ compared to high-$Q$ data.
By contrast, most $Q$-associated variants reported in the GWAS catalogue had weak bias (See Figure \ref{Pub_30}), with three exceptions.
One study that reports associations with seven highly biased $Q$-associated variants considered copy number of Epstein-Barr virus (EBV) sequence in the 1kGP as a phenotype \citep{Mandage2017}. 
The authors noted that these seven variants were not flanked by LD peaks and were correctly removed from downstream analyses -- though still included in the NHGRI-EBI Catalog \citep{Mandage2017}.
It is plausible that the EBV copy number phenotype is sensitive to mapping and confounded by the same technical artefacts that lead to biased SNP calling.

@@

\section{Author's Note}
\todo{While reviewing our manuscript for publication, in response to our pre-print, the New York Genome Center announced that they had resequenced the original 1kGP dataset in a yet to be published dataset.}

@@

\section{Acknowledgments}
We would like to thank Kelly Harris for sharing her mutation spectrum scripts.
We would also like to thank the members of the Gravel lab for their help with coding and useful discussions.

>>>>

\begin{figure}
\scalebox{0.85}{\includegraphics[width=\hsize,keepaspectratio]{../Figures/Figure1.jpg}}
\caption{
Suspicious mutations carried by individuals with low quality data have distinct mutational profiles, reproduce poorly across studies, and are distributed across the genome.
\textbf{A} 
Genome wide association of the average quality of mapped bases $Q$ for the 104 Japanese individuals included in the 1000 Genomes Project. 
This GWAS identified $1034$ SNPs associated to the average $Q$ of SNPs mapped for an individual, with  $p < 10^{-6}$
($587$ SNPs had  $p < 10^{-8}.$)
\textbf{B} 
Joint frequency spectrum plot of the Japanese from the 1000 Genomes Project and a more recent Japanese dataset from Nagahama.
The plot is zoomed in on frequencies below 0.5 for clarity.
The size of blue dots are proportional to the number of variants in a given frequency bin that associate with $Q$ in the JPT. 
\textbf{C} 
Mutation spectrum of the 1034 variants that associated with $Q$ in the JPT($p < 10^{-6}$), compared to the expectation from the distribution of all SNPs.
The majority of the variants with significant associations to $Q$ have the *AC${\rightarrow}$*CC mutational pattern. 
There is also an enrichment in GA*${\rightarrow}$GG* and GC*${\rightarrow}$GG* mutations. 
These three enrichments can be summarized as G**${\rightarrow}$GG*. 
Stars ( * ) indicate a significant deviation from the expected mutational spectrum from all SNPs.
}
 \label{SFS}
\end{figure}

@@

These associations are not necessarily incorrect -- a weak but significant bias in imputation may still result in a correct associations.
To distinguish between variants with weak but significant association with $Q$ from variants with strong biases, we distinguished between variants where the allele frequency difference between individuals with low- and high-$Q$ is larger than a factor of two (which naturally separates two clusters of variants on Figure \ref{90HanSFS_full}).
The majority (92.7\%) of the $Q$-associated variants are strongly biased in that they are more than twice as frequent in individuals with low-$Q$ compared to high-$Q$ data.
By contrast, most $Q$-associated variants reported in the GWAS catalogue had weak bias (See Figure \ref{Pub_30}), with three exceptions.
One study that reports associations with seven highly biased $Q$-associated variants considered copy number of Epstein-Barr virus (EBV) sequence in the 1kGP as a phenotype \citep{Mandage2017}. 
These seven variants were not flanked by LD peaks and were correctly removed from downstream analyses -- though still included in the NHGRI-EBI Catalog \citep{Mandage2017}.
It is plausible that the EBV copy number phenotype is sensitive to mapping and confounded by the same technical artefacts that lead to biased SNP calling.

@@

\section{Author's Note}
\todo{While revising our manuscript for publication, in response to our pre-print, the New York Genome Center announced that they had resequenced the original 1kGP dataset in a yet to be published dataset.}

@@

\section{Acknowledgments}
We would like to thank Kelly Harris for sharing her mutation spectrum scripts.
\todo{We would like to thank Rajendra Mandage, Gabriel Santpere and Arcadi Navarro for sharing more details relating to the results from their publication.}
We would also like to thank the members of the Gravel lab for their help with coding and useful discussions.
